---
title: 'Election Blog #3: Polling Predictions'
author: Shivali Korgaonkar
date: '2024-09-22'
slug: election-blog-3
categories: []
tags: []
---

```{r libraries, include= FALSE}

library(car)
library(caret)
library(CVXR)
library(glmnet)
library(tidyverse)
library(dplyr)
library(lubridate)

# Load in presidential poll ratings
ratings_2016 <- read_csv('~/Downloads/president_polls_2016.csv')
ratings_2020 <- read_csv('~/Downloads/president_polls_2020.csv')
ratings_2024 <- read_csv('~/Downloads/president_polls_2024.csv')

# Personal Blog Theme
my_blog_theme <- function() {
  theme(
    panel.border = element_blank(),
    panel.background = element_rect(fill = "gray92"),
    plot.title = element_text(size = 18, hjust = 0.8, color = "gray16", face = "bold", family = "Times New Roman"), 
    plot.subtitle = element_text(size = 14, color = "black", hjust = 0.5, family = "Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 10, family = "Times New Roman"),
    axis.line = element_line(colour = "black"),
    axis.title = element_text(size = 12, family = "Times New Roman"),
    legend.position = "right",
    legend.text = element_text(size = 12))
}

```

**Poll Ratings Grade Distribution**

```{r include=FALSE}
##Extension 1
#Consider 2016-2024 pollster ratings from FiveThirtyEight. (1.) How much variation is there in pollster quality?

# 2016 Histogram of Poll Ratings
ratings_2016$grade <- factor(ratings_2016$grade, levels = c("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D", "NA"))
bar_2016_pollratings <- ratings_2016|>
  ggplot(mapping = aes(x = grade)) + 
  geom_bar(fill = "plum4") +
  labs(title = "2016 Poll Ratings Grade Distribution",
       x = "Grade",
       y = "Count") +
  my_blog_theme()
bar_2016_pollratings
```

Among the 2016 presidential election polls provided this week, letter grades were designated to disclose their reliability. The most common poll grades were A-, B, and C-, respectively.

```{r include=FALSE}
# 2020 Histogram of Poll Ratings
ratings_2020 <- ratings_2020 |>
  rename(grade = fte_grade)

ratings_2020$grade <- factor(ratings_2020$grade, levels = c("A+", "A", "A-", "A/B", "B+", "B", "B-", "B/C", "C+", "C", "C-", "C/D", "F", "NA"))
bar_2020_pollratings <- ratings_2020|>
  ggplot(mapping = aes(x = grade)) + 
  geom_bar(fill = "sandybrown") +
  labs(title = "2020 Poll Ratings Grade Distribution",
       x = "Grade",
       y = "Count") +
  my_blog_theme()
bar_2020_pollratings
```
In 2020, the most common letter grades for the polls provided was overwhelmingly C, with a distant second of B/C. The letters grades are significantly lower graded than in 2016, and the distribution is left skewed. 

```{r include=FALSE}
# 2024 Histogram of Poll Ratings
hist_2024_pollratings <- ratings_2024|>
  ggplot(mapping = aes(x = numeric_grade)) + 
  geom_histogram(bins = 15, fill = "darkseagreen", color = "snow3") +
  scale_x_reverse() +
  labs(title = "2024 Poll Ratings Grade Distribution",
       x = "Grade",
       y = "Count") +
  my_blog_theme()
hist_2024_pollratings
```
The grading scheme changes for 2024 to a 1-3 scale with 3 being the highest rating. There's a heavy emphasis on the higher score between the 2 and 3 range.

**Polling Averages**

```{r 2016, include=FALSE}
# (2.) Using tools and knowledge you have gained so far, build a model (or ensemble model) that uses individual polls from 2016 (president_polls_2016.csv), 2020 (...2020.csv), and 2024 (...2024.csv). How does your model compare to the models this week in lab?

# Finding the end date average in 2016 polls
election_day_2016 <- as.Date("2016-11-08")

nat_ratings_2016 <- ratings_2016 |>
  mutate(enddate = as.Date(enddate, format = "%m/%d/%Y")) |>
  filter(state == "U.S.") |>
  group_by(enddate) |>
  summarize(mean(adjpoll_clinton),
            mean(adjpoll_trump)) |>
  rename(avg_clinton = "mean(adjpoll_clinton)",
         avg_trump = "mean(adjpoll_trump)")

nat_ratings_2016 <- nat_ratings_2016 |>
  pivot_longer(cols = c(avg_clinton, avg_trump),
               names_to = "candidate", values_to = "average") |>
  mutate(candidate = if_else(candidate == "avg_clinton", "Hillary Clinton", "Donald Trump")) |>
  rename(pct = average) |>
  mutate(days_until_election = as.numeric(difftime(election_day_2016, enddate, units = "days")),
         weeks_until_election = days_until_election / 7)

nat_pollagg_2016_plot <- nat_ratings_2016 |>
  ggplot(aes(x=enddate, y = pct, color = candidate)) +
  geom_point(alpha = 0.25, size = 1.5) +
  geom_smooth(se = FALSE, span = .08) +
  scale_x_date(date_labels = "%b %d") +
  scale_color_manual(values = c("firebrick3", "dodgerblue3")) +
  labs(x = "Date",
       y = "Average Poll Approval",
       title = "Polling Averages by Date, 2016") +
  my_blog_theme()
nat_pollagg_2016_plot


```
The graph above tracks poll approval ratings for Trump and Clinton for about one year before election day. There a few notable trends in this chart. For starters, there is a drastic dip in December 2015 for Trump. This sink in polls can likely be attributed to Trump calling for a complete shutdown on Muslim people entering the United States seeking refuge or citizenship. Additionally, before February 2016, the candidates followed each other in poll approval, meaning they rose and sunk together. Afterwards, their votes began to diverge, so when Clinton did well, Trump did not, and vice versa. This is because the party primaries began in 2016, so both parties had great division between nominees. The Republicans lost a more moderate Ted Cruz and Marco Rubio, while the democrats lost a more progressive Bernie Sanders.

```{r 2020, include=FALSE}

# End date average
election_day_2020 <- as.Date("2020-11-03")
nat_ratings_2020 <- ratings_2020 |>
  mutate(enddate = as.Date(end_date, format = "%m/%d/%y")) |>
  filter((is.na(state))) |>
  filter(answer %in% c("Trump", "Biden")) |>
  filter(enddate >= as.Date("2019-11-01")) |>
  group_by(enddate, candidate_name) |>
  summarize(mean(pct)) |>
  rename(pct = "mean(pct)")|>
  mutate(days_until_election = as.numeric(difftime(election_day_2020, enddate, units = "days")),
         weeks_until_election = days_until_election / 7) |>
  rename(candidate = candidate_name)

  
# Each enddate average
nat_pollagg_2020_plot <- nat_ratings_2020 |>
  ggplot(aes(x=enddate, y = pct, color = candidate)) +
  geom_point(alpha = 0.25, size = 1.5) + 
  geom_smooth(se = FALSE, span = .08) + 
  scale_x_date(date_labels = "%b %d") + 
  scale_color_manual(values = c("firebrick3", "dodgerblue3")) +
  labs(x = "Date",
       y = "Average Poll Approval", 
       title = "Polling Averages by Date, 2020") + 
  my_blog_theme()
nat_pollagg_2020_plot

```
The 2020 poll approval chart does not display extreme dips and peaks like in 2016. There is a similar trend that follows before and after the primaries, but not as drastically, since there were no major candidates running against Trump in 2020. The polls find their predictions converge near election day, unlike in 2016 as well. Between April and July, Trump struggled to consistently raise his approval, as the COVID-19 pandemic was endangering millions of Americans everyday. George Floyd's death in May also sparked the beginning of the Black Lives Matter movement, which the former President had been unable to empathize with or address, explaining his low rating.

```{r 2024, include=FALSE}

# Each enddate average
election_day_2024 <- as.Date("2024-11-05")
nat_ratings_2024 <- ratings_2024 |>
  mutate(enddate = as.Date(end_date, format = "%m/%d/%y")) |>
  filter((is.na(state))) |>
  filter(answer %in% c("Trump", "Harris")) |>
  filter(enddate >= as.Date("2023-11-01")) |>
  group_by(enddate, candidate_name) |>
  summarize(mean(pct)) |>
  rename(pct = "mean(pct)") |>
  mutate(days_until_election = as.numeric(difftime(election_day_2024, enddate, units = "days")),
         weeks_until_election = days_until_election / 7) |>
  rename(candidate = candidate_name)

nat_pollagg_2024_plot <- nat_ratings_2024 |>
  ggplot(aes(x=enddate, y = pct, color = candidate)) +
  geom_point(alpha = 0.25, size = 1.5) + 
  geom_smooth(se = FALSE, span = .08) + 
  scale_x_date(date_labels = "%b %d") + 
  scale_color_manual(values = c("firebrick3", "dodgerblue3")) +
  labs(x = "Date",
       y = "Average Poll Approval", 
       title = "Polling Averages by Date, 2024") + 
  ylim(35,50)+
  my_blog_theme()
nat_pollagg_2024_plot

```
The chaos of 2024's polls represents the chaos that has undergone politics over the past year. The blue line represents President Biden's polls until July 21, 2024 when he drops out of the race and endorses Kamala Harris. Trump and Biden generally matched each other, staying between the 40-45% approval range. However, after a disastrous first debate at the end of June, Biden's polls sink to 35%. Immediately after he drops out in July, the Democratic Party jolts upwards and continues to have a comfortable lead above Trump, despite the age of Harris' campaign, especially following the Democratic National Convention.

```{r include=FALSE}
####----------------------------------------------------------#
#### Regularized regression with polling data.
####----------------------------------------------------------#
# Read election results data. 
d_vote <- read_csv("~/Downloads/popvote_1948-2020.csv")
d_vote$party[d_vote$party == "democrat"] <- "DEM"
d_vote$party[d_vote$party == "republican"] <- "REP"

# Combine all national poll data (2016-2024)
combined_polls <- nat_ratings_2016 |>
  full_join(nat_ratings_2020) |>
  full_join(nat_ratings_2024)
```


```{r include=FALSE}
####----------------------------------------------------------#
#### Model ensembling.
####----------------------------------------------------------#

# Estimate models using polls alone, fundamentals alone, and combined fundamentals and polls.
# Read economic data.
d_econ <- read_csv("~/Downloads/fred_econ.csv") |>
  filter(quarter == 2)
# 
# # Combine datasets and create vote lags. 
# d_combined <- d_econ |> 
#   left_join(d_poll_weeks, by = "year") |> 
#   filter(year %in% c(unique(d_vote$year), 2024)) |> 
#   group_by(party) |> 
#   mutate(pv2p_lag1 = lag(pv2p, 1), 
#          pv2p_lag2 = lag(pv2p, 2)) |> 
#   ungroup() |> 
#   mutate(gdp_growth_x_incumbent = GDP_growth_quarterly * incumbent, 
#          rdpi_growth_quarterly = RDPI_growth_quarterly * incumbent,
#          cpi_x_incumbent = CPI * incumbent,
#          unemployment_x_incumbent = unemployment * incumbent,
#          sp500_x_incumbent = sp500_close * incumbent) # Generate interaction effects.
```


**Sources**

(https://muslimadvocates.org/wp-content/uploads/2019/06/Timeline-of-Record-of-Bigotry.pdf).